Converging High-Throughput and High-Performance Computing Paradigms: \\ A Case Study of Using Abstractions to Bridge the Divide

### 2017-03-18 overall.tex
Section V ready for review. Iteration, figures, captions and addressed comments

Figure~\ref{fig:core-hours-utilization} shows the number of core-hours used by
ATLAS on Titan during the experiment time window. ATLAS consumed a total of
73.8M core-hours, for an average of $\approx$ 7M core-hours a month, with a
minimum of 3.3M core-hours in April 2016 and a maximum 14.8M core-hours in
February 2017.

@@

\begin{figure}[htp]
\includegraphics[clip,width=\columnwidth]{figures/cpu_hours.png}
\caption{Caption.}
\label{fig:core-hours-utilization}
\end{figure}

Figure~\ref{fig:backfill-utilization} shows backfill utilization during the
experiment time window.\jhanote{Neither figure show efficiency. Both show
utilization.} Efficiency is defined as the fraction of Titan’s cores available
via backfill that was utilized by the PanDA Brokers. \jhanote{I would plot the
efficiency on the Y2 axis for both.} The brokers reached 18\% average
efficiency, with a minimum 8.9\% efficiency on April 2016 and a maximum 33.5\%
efficiency on February 2017. The number of total backfill cores available was
38.1M in April 2016, and 33.1M in February 2017. This shows that the measured
difference in efficiency did not depend on a comparable difference in total
backfill availability.

\begin{figure}[htp]
\includegraphics[clip,width=\columnwidth]{figures/backfill_consumption.png}
\caption{Caption. \jhanote{I would suggest similar style of X-axis as previous figure. Difficult to parse which is April if we are going to call out a specific month. Also consistency in style is generally good to keep cognitive burden low.}}
\label{fig:backfill-utilization}
\end{figure}

Figure~\ref{fig:hpc-workload-utilization} shows that during the experiment time
window, about 2.25M detector simulation jobs were completed on Titan, for a
total of 225M events processed. This is equivalent to 0.9\% of all the 250M
detector simulations performed by ATLAS in the same period of time, and 3.5\% of
the 6.6B events processed by those jobs. Comparatively, Titan contributed 3.9\%
of the total of around 200K cores available to ATLAS on Grid, Cloud, and HPC
infrastructures together. These figures confirms the relevance of
supercomputers' resource contribution to the LHC Run 3, especially when
accounting for the amount of unused backfill availability and the rate of
improvement of PanDA efficiency.

\begin{figure}[htp]
    \includegraphics[clip,width=\columnwidth]{figures/cpu_hours.png}
\caption{\mtnote{Placeholder for the diagram/data I asked to Danila and
Sergey. Consider a table instead of a figure.}}
\label{fig:hpc-workload-utilization}
\end{figure}

@@

Every detector simulation executed on Titan process 100 events. This number of
events is consistent with the physics of the use case and with the average
duration of backfill availability. The duration of a detector simulation is a
function of the number of events: the more events, the longer the simulation.
Not all events take the same time to be simulated:
Figure~\ref{fig:100event-distrib} shows a distribution of the simulation time of
1 event from $\approx$ 2 to $\approx$ 40 minutes, with a mean of $\approx$ 14
minutes. Considering that each worker node process up to 16 events concurrently,
100 events takes around 2 hours to process.

\begin{figure}[htp]
    \includegraphics[clip,width=\columnwidth]{figures/tx_simulation_distribution.pdf}
\caption{Distribution of the time taken by a Geant4 detector simulation to simulate one event. 3000 Events; 30 Titan worker nodes; 16 simulations per node; 100 events per node. Average time per event $\approx$ 14 min. Broad distribution from $\approx$ 2 to $\approx$ 40 minutes.}
\label{fig:100event-distrib}
\end{figure}

Figure~\ref{fig:backfill-distrib} shows the distribution of backfill
availability on Titan as a function of number of nodes and the time of their
availability (i.e., walltime). The mean number of nodes over the experiment
window time was $\approx$ 267, and their mean walltime was 1445 minutes.
Detector simulations of 100 events, enable to use down to 1 node for 1/4 of the
mean walltime of backfill availability. As such, it offers a good compromise for
the PanDA Broker efficiency.

\begin{figure}[htp]
    \includegraphics[clip,width=\columnwidth]{figures/avail_time_vs_nodes_Oct_28_Nov_3_2015.png}
\caption{\mtnote{Waiting to know whether we have data for the experiment time window.}}
\label{fig:backfill-distrib}
\end{figure}

@@

\begin{itemize}
    \item problem with lustre addressed with ramdisk (done)
    \item Competition for cache in AMD processors. (done)
    \item Comparison hardware performance Titan/Grid (data acquired)
    \item Summary overall figure of performance analogous to those produced with NGE. Diagram as discussed in F2F meeting at Rutgers. (waiting)
\end{itemize}

@@

After careful characterization of the impact of ATLAS jobs on Lustre, the
performance degradation was addressed by moving the AthenaMP distribution to a
read-only NFS directory, shared among DTNs and worker nodes. This eliminated the
problem of metadata contention, improving metadata read performance from
$\approx$ 6,300 seconds on Lustre to $\approx$ 1,500 seconds on NFS. Further, at
configuration time, the input files of each detector simulation job were stored
to a ramdisk on the work nodes. This offered a marked improvement of reading
time: from 1,320 seconds on Lustre to 40 seconds on NFS.

@@

The compute performance of Titan for detector simulation jobs is, on average,
lower that that of LCWG. Figure~\ref{fig:comparison-titan-grid} shows the
difference in execution time of the same type of simulation between Titan and
\ldots\mtnote{Add the name of the grid sites here}. This is not an issue with
the design of PanDA Brokers and, more in general, with the migration of part of
the ATLAS workload to supercomputers: Titan has been active since 2012 and more
modern CPUs are expected to offer better performance. A pilot project is already
in place to test PanDA Broker on Summit, the next generation supercomputer that
will become active in 2018 at OLCF.

\begin{figure}[htp]
    \includegraphics[clip,width=\columnwidth]{figures/tx8_tx16_comparison.pdf}
    \caption{}
\label{fig:comparison-titan-grid}
\end{figure}

@@

Figure~\ref{fig:failures-titan} shows a breakdown of the type of failures
recorded by the PanDA Brokers between January 2016 and February 2017. The
initial preponderance of unknown failures was progressively addressed by a more
accurate failure model, specifically tailored to the PanDA Broker and Titan. It
also shows the progressive improvement of the broker's reliability: In Februrary
2017, most of the failures were produced by the initialization and configuration
phases of AthenaMP or by failed executions of the payload.\mtnote{If we want to
keep this diagram, we will need to plot it from the CSV to make it readable.}

\begin{figure}[htp]
    \includegraphics[clip,width=\columnwidth]{figures/failures_titan.png}
\caption{Caption.}
\label{fig:failures-titan}
\end{figure}

>>>>

Figure~\ref{fig:backfill-utilization} (gray bars) shows the number of core-hours
used by ATLAS on Titan during the experiment time window. ATLAS consumed a total
of 73.8M core-hours, for an average of $\approx$7M core-hours a month, with a
minimum of 3.3M core-hours in April 2016 and a maximum 14.8M core-hours in
February 2017.

@@

% \begin{figure}[htp]
% \includegraphics[clip,width=\columnwidth]{figures/cpu_hours.png}
% \caption{Caption.}
% \label{fig:core-hours-utilization}
% \end{figure}

\begin{figure}[htp]
    \includegraphics[clip,width=\columnwidth]{figures/backfill_consumption.pdf}
    \caption{Gray bars: Number of unused cores of Titan's backfill availability;
    green bars: Number of cores of backfill availability used by ATLAS via PanDA
    Brokers on Titan; red labels: Efficiency of PanDA Brokers on Titan defined
    as percentage of backfill availability used every month of the experiment
    window.}
\label{fig:backfill-utilization}
\end{figure}

Figure~\ref{fig:backfill-utilization} (green bars) shows backfill utilization
during the experiment time window. Efficiency
(Figure~\ref{fig:backfill-utilization}, red labels) is defined as the fraction
of Titan’s cores available via backfill that was utilized by the PanDA Brokers.
The brokers reached 18\% average efficiency, with a minimum 8.9\% efficiency on
April 2016 and a maximum 33.5\% efficiency on February 2017. The number of total
backfill cores available was 38.1M in April 2016, and 33.1M in February 2017.
This shows that the measured difference in efficiency did not depend on a
comparable difference in total backfill availability.

\jhanote{I would plot the efficiency on the Y2 axis for both
[diagrams].}\mtnote{Done but then the first diagram ended up plotting a subset
of the data of the second diagram. I used only the second diagram.} \jhanote{I
would suggest similar style of X-axis as previous figure. Difficult to parse
which is April if we are going to call out a specific month. Also consistency in
style is generally good to keep cognitive burden low.}\mtnote{Done.}

% Figure~\ref{fig:hpc-workload-utilization} shows that d

During the experiment time window, about 2.25M detector simulation jobs were
completed on Titan, for a total of 225M events processed. This is equivalent to
0.9\% of all the 250M detector simulations performed by ATLAS in the same period
of time, and 3.5\% of the 6.6B events processed by those jobs. Comparatively,
Titan contributed 3.9\% of the total of around 200K cores available to ATLAS on
Grid, Cloud, and HPC infrastructures together. These figures confirms the
relevance of supercomputers' resource contribution to the LHC Run 3, especially
when accounting for the amount of unused backfill availability and the rate of
improvement of PanDA efficiency.

% \begin{figure}[htp]
%     \includegraphics[clip,width=\columnwidth]{figures/cpu_hours.png}
% \caption{\mtnote{Placeholder for the diagram/data I asked to Danila and
% Sergey. Consider a table instead of a figure.}}
% \label{fig:hpc-workload-utilization}
% \end{figure}

@@

Every detector simulation executed on Titan process 100 events. This number of
events is consistent with the physics of the use case and with the average
duration of backfill availability. The duration of a detector simulation is a
function of the number of events: the more events, the longer the simulation.
Not all events take the same time to be simulated:
Figure~\ref{fig:100event-distrib} shows a distribution of the simulation time of
1 event from $\approx$2 to $\approx$40 minutes, with a mean of $\approx$14
minutes. Considering that each worker node process up to 16 events concurrently,
100 events takes around 2 hours to process.

\begin{figure}[htp]
    \includegraphics[clip,width=\columnwidth]{figures/tx_simulation_distribution.pdf}
    \caption{Distribution of the time taken by a Geant4 detector simulation to
    simulate one event. 3000 Events; 30 Titan worker nodes; 16 simulations per
    node; 100 events per node. Average time per event $\approx$14 min. Broad
    distribution from $\approx$2 to $\approx$40 minutes.}
\label{fig:100event-distrib}
\end{figure}

Figure~\ref{fig:backfill-distrib} shows the distribution of backfill
availability on Titan as a function of number of nodes and the time of their
availability (i.e., walltime). We recorded these data by polling Titan's Moab
scheduler at regular intervals between Oct 28 and Nov 3 2015, while developing
PanDA Broker. The mean number of nodes was 267.1, and their mean walltime was
1445 minutes. Detector simulations of 100 events, enable to use down to 1 node
for 1/4 of the mean walltime of backfill availability. As such, it offers a good
compromise for PanDA Broker efficiency.

\begin{figure}[htp]
    \includegraphics[clip,width=\columnwidth]{figures/avail_time_vs_nodes_Oct_28_Nov_3_2015.pdf}
    \caption{Backfill availability measures as walltime vs number of cores
    between Oct 28 and Nov 2015. Availability was measured 5312 times: mean
    number of work nodes available 267.1; mean walltime available 1445
    minutes.\mtnote{Waiting to know whether we have data for the experiment
    time window.}}
\label{fig:backfill-distrib}
\end{figure}

@@

% \begin{itemize}
%     \item problem with lustre addressed with ramdisk (done)
%     \item Competition for cache in AMD processors. (done)
%     \item Comparison hardware performance Titan/Grid (data acquired)
%     \item Summary overall figure of performance analogous to those produced
%     with NGE. Diagram as discussed in F2F meeting at Rutgers. (waiting)
% \end{itemize}

@@

After careful characterization of the impact of ATLAS jobs on Lustre, the
performance degradation was addressed by moving the AthenaMP distribution to a
read-only NFS directory, shared among DTNs and worker nodes. This eliminated the
problem of metadata contention, improving metadata read performance from
$\approx$6,300 seconds on Lustre to $\approx$1,500 seconds on NFS. Further, at
configuration time, the input files of each detector simulation job were stored
to a ramdisk on the work nodes. This offered a marked improvement of reading
time: from 1,320 seconds on Lustre to 40 seconds on NFS.

@@

The performance analysis of Titan's AMD CPUs for detector simulations helps also
to compare Titan and Grid site performances. Usually, Grid sites exposes
resources with heterogeneous CPU architectures and a maximum of 8 (virtual)
cores per worker node, while Titan's offer an homogeneous 16 cores architecture.
We used the rate of events processes per minute as a measure of the efficiency
of executing the same detector simulation on both Titan or Grid sites.
Figure~\ref{fig:comparison-titan-grid} compares the efficiencies of Titan to the
BNL and SIGNET Grid sites, normalized for 8 cores. Effective performance
per-core at Titan is $\approx$0.57 event per minute, roughly 1/2 of BNL and 1/3
of SIGNET performances.

\begin{figure}[htp]
    \includegraphics[clip,width=\columnwidth]{figures/titan_grid_event_rate.pdf}
    \caption{Comparison of the event rate for finished jobs at Titan (h1), BNL
    (h2) and GRIDNET (h3). Titan histogram shows 1/2 of the measured values to
    account for the core difference between Titan (16) and the Grid sites (8).
    Performance per core: BNL 1.126; SIGNET 0.885; Titan 0.577.}
\label{fig:comparison-titan-grid}
\end{figure}

The differences in performance between Titan and the BNL and SIGNET Grid sites
are due to the FP scheduler competition and the availability of newer
processors. The CPUs at the Grid sites have one FP scheduler per core and are on
average newer than the CPU of Titan. The heterogeneity of the Grid sites' CPUs
explain the higher performance variance compared to the performance consistency
measured on Titan.

\mtnote{Figure summarizing job performance on Titan analogous to those produced
with NGE, as discussed in F2F meeting at Rutgers?}

@@

Figure~\ref{fig:failures-titan} shows a breakdown of the type of failures
recorded by the PanDA Brokers between January 2016 and February 2017. The
initial preponderance of unknown failures was progressively addressed by a more
accurate failure model, specifically tailored to the PanDA Broker and Titan.
Figure~\ref{fig:failures-titan} also shows the progressive improvement of the
broker's reliability: In Februrary 2017, errors related to input files and
initialization were almost eliminated and execution errors were comparable to
those recorded in January 2017 even if almost twice as many simulations were
performed (Figure~\ref{fig:backfill-utilization}). The increased amount of
unknown failure in February indicates that furter work is needed to improve the
PanDA Broker failure model.

\begin{figure}[htp]
    \includegraphics[clip,width=\columnwidth]{figures/failures_titan.pdf}
    \caption{PanDA Broker failures during the experiment time window. Failures
    were recorded with 52 exit codes, here aggregated into 6 categories. The
    number of failures should be compared to the number of simulations executed
    in each month as plotted in Figure~\ref{fig:backfill-utilization}.
    Improvements in the failure model of PanDA Broker shows a decrease in
    unknown failure and a the progressive addressing of other categories of
    failure.}
\label{fig:failures-titan}
\end{figure}
